Future challenges for Research Software Engineering

### 2025-04-17 future_challenges_for_RSEng.tex

\subsection{Ethics and social consideration}
FG-NOTE: Given what ChatGPT has put in there, I wonder if we need to separate these two items
MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?

- Military or other sensitive applications:
  - RSEs (like physicists, mathematicians, etc.) might face ethical dilemmas if their work can be repurposed for weaponry.
  - Importance of having personal “red lines” and awareness of how software can be used.

- Corporate dominance:
  - AI and coding platforms might be increasingly controlled by a small number of large corporations.
  - RSE community might need to develop or advocate for open-source alternatives.

>>>>

\subsection{Ethics and social consideration}
The ethics and values of the Research Software Engineer are grounded in their practice.
They commit to the health, safety and welfare of the public and act in the interest of society, their employer and their clients \cite{Goth2023}.
Sometimes these values are in direct conflict with the aims of their institution.
For example, their PI might insist on quickly developing some software without adhering to current best practices.
Another example might be that they are developing code that can also be used for military purposes, such as numerical solvers or computer vision systems.
Like engineers, RSEs take the responsibility for the software they develop.
RSEs need to be aware of these issues and develop their own ``red lines''.

RSEs are publicly funded.
Similarly to how data collected by scientists should adhere to the FAIR principles \cite{FAIR}, research software developed by RSEs should adhere to the FAIR4RS principles~\cite{FAIR4RS}.

A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines.

MM-NOTE: (Flo believes it was from Michele) Are we assuming that progress will bring prosperity? Will we remain constrained by needing to produce products/artifacts?

  - Importance of having personal “red lines” and awareness of how software can be used.

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay have started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious. Kay is working in a project that develops
software for cheap water monitoring devices in the global south. Kim is currently designing the software service stack
at her university, and increasingly struggles to find compelling Open Source alternatives that let's them keep their data
to software that is backed by large corporate entities. So far she still has the support of her university board, since it values
strategic autonomy.
Her son Kyle on the other hand started out with a degree in Physics and then took one of the newly created RSE Master courses,
but quickly decided that work in academia with its time limited contracts is not attractive.
Governments around the world are struggling to reposition themselves in this emerging deglobalized world with its new blocks
and therefore have deprioritized publicly funded research. Kyle decided to go to a defence company since they offer a
more secure employment in this environment.
\end{story}

\subsection{What are new tasks of RTPs?}
 FG-NOTE: RSEs are one type of specialization of RTPs in science. Do we see any other?

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
